---
title: "Leah_EDA"
author: "Leah Ekblad"
date: "2024-09-15"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_location: left
    toc_title: "Contents"
execute:
  include: true
  eval: true    
  warning: false
  message: false
---

# Introduction
Swire Coca-Cola (SCCU) is optimizing its logistics strategy by transitioning low-volume customers to an Alternate Route to Market (ARTM) using third-party (white truck) deliveries, while prioritizing high-volume customers for direct (red truck) deliveries. The goal is to determine an optimal sales volume threshold for transitioning customers, identify high-growth-potential accounts, and integrate predictive insights into SCCU’s routing strategy.

This analysis is supported by four datasets: Customer Profile, Transactional Data, Delivery Cost Data, and ZIP Mapping Data. These datasets provide insights into customer purchasing behavior, delivery costs, and geographic distribution, which will inform data-driven logistics decisions.

## Business Problem
SCCU faces challenges in balancing cost efficiency and customer service in its delivery operations. The key problems include:

Optimizing Route-to-Market (RTM) Strategy – Determining which customers should be shifted to ARTM to reduce operational costs while maintaining service quality.
Identifying High-Potential Customers – Ensuring that high-growth-potential customers are retained within the direct delivery network (red truck) to foster long-term profitability.
Cost Analysis & Threshold Definition – Establishing clear volume-based thresholds for transitioning customers while considering delivery cost structures.
This analysis aims to provide data-driven recommendations to refine SCCU’s logistics model.

# Description of the Data
1. Customer Profile Data
30,478 customers with details on their ordering habits, trade channels, and onboarding dates.
Key fields: customer_number, trade_channel, on_boarding_date, first_delivery_date, local_market_partner.
2. Transactional Data
Over 1 million records tracking weekly transactions, including order volume and delivery fulfillment.
Key fields: customer_number, transaction_date, ordered_cases, delivered_cases, ordered_gallons, delivered_gallons.
3. Delivery Cost Data
Contains cost structures for different cold drink channels, showing delivery costs per unit for various product types.
Key fields: cold_drink_channel, applicable_to, median_delivery_cost, cost_type, min_volume, max_volume.
4. ZIP Mapping Data
1,801 ZIP codes linked to geographic information for understanding regional distribution.
Key fields: zip_code, full_address.

# Initial Questions
Threshold Identification: What is the optimal sales volume threshold to determine whether a customer should receive red truck vs. white truck delivery?
Growth Potential: Can we identify customers with potential for increased purchases who should remain in the direct delivery network?
Order & Delivery Efficiency: How often do customers receive full vs. partial order fulfillment, and does this vary by RTM type?
Delivery Cost Analysis: How do median delivery costs vary across different customer segments and volume ranges?
Market Segmentation: Are there specific trade channels or geographic regions where ARTM or direct delivery would be more effective?
The answers to these questions will guide SCCU in optimizing its logistics strategy, reducing costs, and improving service levels.


```{r setup, include=FALSE}
#| include: false

# Packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(caret)
library(janitor)

# Data
transact_data <- read_csv("transactional_data.csv")
customer_profile <- read_csv("customer_profile.csv")
customer_address_and_zip_mapping <- read_csv("customer_address_and_zip_mapping.csv")
cost_data <- read_csv("delivery_cost_data.xlsx")
```

## customer_address_and_zip_mapping
### Cleaning
- Separated out address into multiple columns
```{r}
# Clean customer address and zip mapping data 

# Use strsplit to split the full_address by commas
clean_customer_address_and_zip_mapping <- strsplit(customer_address_and_zip_mapping$'full address', ",")

# Convert the split data into a data frame
# Use sapply to apply strsplit to the entire column and then transpose the result using t()
clean_customer_address_and_zip_mapping <- as.data.frame(t(sapply(clean_customer_address_and_zip_mapping, c)))

# Assign column names
colnames(clean_customer_address_and_zip_mapping) <- c("Zip Code", "City", "State", "StateAbbr", "County", "Region", "Latitude", "Longitude")
```

## transact_data
### Cleaing
- Converted date fields to the Date format
- Handled missing values: 
  - This resulted in no rows omitted because only transaction date has NAs - I dont think we want to remove those because it still has all of the other data as well as the week of the year

### Feature Engineering
- Broke out date into Month and day, year and week were already created
- Aggregated by customer for total annual orders
- Order fufillment ratio
  - There are a lot of NAs
  - This feature measures how much of the ordered quantity was actually delivered. It's an important metric for understanding order fulfillment efficiency and customer satisfaction
  - Interpretation: A ratio of 1 means full fulfillment, while less than 1 indicates partial fulfillment
- Order Backlog: This metric indicates if there are any outstanding orders that were not delivered. It's useful to identify customers that are frequently unable to receive their full order quantity.
  - Interpretation: A positive value indicates unfulfilled orders, while a value of 0 indicates complete fulfillment.
- Creating a 'Season' feature based on the month of the transaction date
```{r}
# Clean transact data

# View the data
summary(transact_data)

# Convert the TRANSACTION_DATE to Date type
transact_data$TRANSACTION_DATE <- as.Date(transact_data$TRANSACTION_DATE)

# Filter out rows with missing values in key columns
# This results in no rows omitted because only transaction date has NAs - I don't think we want to remove those because it still has all of the other data as well as the week of the year
transact_data_clean <- transact_data %>%
  filter(!is.na(ORDERED_CASES) & !is.na(LOADED_CASES) & !is.na(DELIVERED_CASES) & 
         !is.na(ORDERED_GALLONS) & !is.na(LOADED_GALLONS) & !is.na(DELIVERED_GALLONS))

# Feature Engineering

# Create additional columns for Month and day
transact_data_clean$MONTH <- month(transact_data_clean$TRANSACTION_DATE)
transact_data_clean$DAY <- day(transact_data_clean$TRANSACTION_DATE)

# Create a new column for the total ordered volume (cases + gallons) for comparison purposes
transact_data_clean$Total_Ordered_Volume <- transact_data_clean$ORDERED_CASES + transact_data_clean$ORDERED_GALLONS

# Order Fulfillment Ratio
transact_data_clean$Fulfillment_Ratio <- transact_data_clean$DELIVERED_GALLONS / transact_data_clean$ORDERED_GALLONS

# Order Backlog
transact_data_clean$Backlog_Gallons <- transact_data_clean$ORDERED_GALLONS - transact_data_clean$DELIVERED_GALLONS

# Creating a 'Season' feature based on the month of the transaction date
transact_data_clean$Season <- case_when(
  month(transact_data_clean$TRANSACTION_DATE) %in% c(12, 1, 2) ~ "Winter",   # December, January, February
  month(transact_data_clean$TRANSACTION_DATE) %in% c(3, 4, 5) ~ "Spring",   # March, April, May
  month(transact_data_clean$TRANSACTION_DATE) %in% c(6, 7, 8) ~ "Summer",   # June, July, August
  month(transact_data_clean$TRANSACTION_DATE) %in% c(9, 10, 11) ~ "Fall",   # September, October, November
  TRUE ~ "Unknown"  # Default case if something unexpected happens
)

```

## customer_profile
### Cleaning
- Convert date fields to the Date format
- Handle missing values: 
  - For categorical columns, filled missing values with "Unknown" or "Other".
  - For numerical columns, filled with median 

### Feature Engineering
  - Calculate the delay between onboarding and first delivery
  - Customer Tenure (Days, Months, Years)
  - Days to First Delivery: Calculates how long it takes for a customer to receive their first delivery after onboarding.
  - Customer Type Based on First Delivery Speed: Categorizes customers based on how fast they received their first delivery.
```{r}
# Clean customer profile data

# View the data
summary(customer_profile)

# Work with a clean version of the data
clean_customer_profile = customer_profile

# Convert `FIRST_DELIVERY_DATE` and `ON_BOARDING_DATE` to Date format
clean_customer_profile$FIRST_DELIVERY_DATE <- as.Date(clean_customer_profile$FIRST_DELIVERY_DATE, format="%m/%d/%Y")
clean_customer_profile$ON_BOARDING_DATE <- as.Date(clean_customer_profile$ON_BOARDING_DATE, format="%m/%d/%Y")


# Check missing values
colSums(is.na(clean_customer_profile))

# Fill missing categorical values with "Unknown"
clean_customer_profile$FREQUENT_ORDER_TYPE[is.na(clean_customer_profile$FREQUENT_ORDER_TYPE)] <- "Unknown"

# Fill missing numerical values (if applicable) with median
clean_customer_profile$FIRST_DELIVERY_DATE[is.na(clean_customer_profile$FIRST_DELIVERY_DATE)] <- median(clean_customer_profile$FIRST_DELIVERY_DATE, na.rm = TRUE)
clean_customer_profile$ON_BOARDING_DATE[is.na(clean_customer_profile$ON_BOARDING_DATE)] <- median(clean_customer_profile$ON_BOARDING_DATE, na.rm = TRUE)

#Feature Engineering
# Calculate the delay between onboarding and first delivery
clean_customer_profile$DAYS_TO_FIRST_DELIVERY <- as.numeric(clean_customer_profile$FIRST_DELIVERY_DATE - clean_customer_profile$ON_BOARDING_DATE)

# Customer Tenure (Days, Months, Years)
clean_customer_profile$CUSTOMER_TENURE_DAYS <- as.numeric(Sys.Date() - clean_customer_profile$ON_BOARDING_DATE)
clean_customer_profile$CUSTOMER_TENURE_MONTHS <- round(clean_customer_profile$CUSTOMER_TENURE_DAYS / 30, 1)
clean_customer_profile$CUSTOMER_TENURE_YEARS <- round(clean_customer_profile$CUSTOMER_TENURE_DAYS / 365, 1)

# Days to first delivery
clean_customer_profile$DAYS_TO_FIRST_DELIVERY <- as.numeric(clean_customer_profile$FIRST_DELIVERY_DATE - clean_customer_profile$ON_BOARDING_DATE)

# Customer Type Based on First Delivery Speed
clean_customer_profile$DELIVERY_SPEED_CATEGORY <- case_when(
  clean_customer_profile$DAYS_TO_FIRST_DELIVERY <= 7 ~ "Fast (0-7 days)",
  clean_customer_profile$DAYS_TO_FIRST_DELIVERY <= 30 ~ "Moderate (8-30 days)",
  clean_customer_profile$DAYS_TO_FIRST_DELIVERY > 30 ~ "Slow (>30 days)",
  is.na(clean_customer_profile$DAYS_TO_FIRST_DELIVERY) ~ "Unknown"
)
```


